---
output:
  pdf_document:
    includes:
      before_body: title.sty
    toc: true
    toc_depth: 1
    latex_engine: xelatex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pander)
library(dplyr)
```
\newpage
# Exploratory Data Analysis
For decompressing the data, run the code chunks in the Rmd file that deem fit for your situation:

* If you downloaded the data from the official site: Run coded chunks `decompress-data` and `create-dataset`.
* If you want to use the dataset delivered with the project: Run the `decompress-dataset` code chunk.

```{r decompress-data, eval=FALSE, echo=FALSE}
#' Use this code chunk to decompress the data downloaded from the site
#' Use the code chunk after this to create the dataset

## Working directory of current project
setwd("labjournal")
cur.wd <- getwd()

## Set the working directory to the data folder (change if needed)
setwd("./Data")

## --- DO NOT CHANGE THE CODE ---
file.names <- list.files()
sapply(file.names, FUN=function (file.name){
  system(paste("gzip -d", file.name))
})
## --- END FIXED CODE ---

## Change the working directory back to the path of the project
setwd(cur.wd)
```


```{r create-dataset, eval=FALSE, echo=FALSE}
#' Use this code chunk to load the different files into one dataset

setwd("../Data")

## Function for reading in files to one dataset
read_sample <- function(file.name) {
  ## Extract the sample name for naming the column
  sample.name <- strsplit(file.name, ".", fixed = TRUE)
  sample.subject <- strsplit(sample.name[[1]][1], "_", fixed=TRUE)[[1]][2]
  colname <- paste0(sample.subject, "_", sample.name[[1]][2])
  ## Read the data, setting the 'Gene' as row.names (column 1)
  sample <- read.table(file.name, sep="\t", row.names=NULL, header=TRUE,
                       col.names=c("Gene", colname))
  ## Return a subset containing the 'Gene' and sample name columns
  return(sample[c(1, 2)])
}

## Get the filenames of the data
file.names <- list.files()

## Read the FIRST sample
dataset <- read_sample(file.names[1])

## Read the rest of the sample group
for (file.name in file.names[2:48]) {
  sample <- read_sample(file.name)
  dataset <- merge(dataset, sample, by = 1)
}

## Exporting completed dataset to a file
setwd("../labjournal")
write.table(dataset, file="gene_count.txt", sep="\t", row.names=FALSE,
            col.names=TRUE, quote=FALSE)

## Now the dataset is ready to use
```


```{r decompress-dataset, eval=FALSE, echo=FALSE}
#' Decompress the complete dataset
#' Use this chunk if you did not download the data from the site and want to use
#'  the delivered gzipped dataset

system("gzip -d gene_count.txt.gz")
```

After decompressing the data, the data can be read:
```{r load-dataset}
## Read the dataset
dataset <- read.table("gene_count.txt", sep="\t", header=TRUE, row.names=1)

## Sample of the dataset
pander(dataset[0:5, 0:3])
```
